Combined annotation exploration for SCPCP000006

### 2025-01-24 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd

---
title: "Combined annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  predicted.celltype.threshold: 0.85
  cnv_threshold: 0
  testing: 0
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

| first level annotation | second level annotation | selection of the cells | marker genes for validation | CNV validation |
| ---------------------- | ----------------------- | ---------------------- | --------------------------- | --------------- |
| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr  & cnv_score < cnv.thr` | `VWF`| no CNV |
| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr & cnv_score < cnv.thr` | `PTPRC`, `CD163`, `CD68`| no CNV |
| normal                 | kidney                  | `cell_type %in% c("kidney cell","kidney epithelial", "podocyte") & predicted.score > pred.thr & cnv_score < cnv.thr` | `CDH1`, `PODXL`, `LTL`| no CNV |
| normal                 | stroma                  | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr`| `VIM`| no CNV |
| cancer                 | stroma                  | `compartment == "stroma" & cnv_score > cnv.thr` | `VIM`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | blastema                | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr`  | `CITED1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | epithelial              | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr`  | `CDH1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| unknown                | -                       | the rest of the cells  | -  | -|

@@

@@

### Output file

The report will be saved in the `notebook` directory.

@@

We simply checked for each chromosome if the cell `has_cnv_chr`.
Would the cell have more than `cnv_threshold` chromosome with CNV, the global `has_cnv_score` will be TRUE.
Else, the cell will have a `has_cnv_score` set to FALSE.

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(has_cnv_score = rowSums(cell_type_df[, grepl("has_cnv_chr", colnames(cell_type_df))])) |>
  mutate(has_cnv_score = case_when(
    has_cnv_score > params$cnv_threshold ~ "CNV",
    has_cnv_score <= params$cnv_threshold ~ "no CNV"
  ))

table(cell_type_df$has_cnv_score)
```

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
# Define normal cells
# We first pick up the immune and endothelial cells annotated via the label transfer compartments under the condition that the predicted score is above the threshold
cell_type_df <- cell_type_df |>
  mutate(first.level_annotation = case_when(
    # assign normal/cancer based on condition
    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "no CNV" ~ "normal",
    compartment %in% c("endothelium", "immune") & cell_type.score > params$predicted.celltype.threshold ~ "normal",
    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "CNV" ~ "cancer",
    .default = "unknown"
  ))
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = first.level_annotation), shape = 19, size = 1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(second.level_annotation = case_when(
    # assign normal cells based on condition
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "no CNV" ~ "kidney",
    cell_type_df$compartment %in% c("stroma") &
      cell_type_df$has_cnv_score == "no CNV" ~ "normal stroma",
    cell_type_df$compartment %in% c("endothelium") &
      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
        cell_type_df$has_cnv_score == "no CNV") ~ "endothelium",
    cell_type_df$compartment %in% c("immune") &
      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
        cell_type_df$has_cnv_score == "no CNV") ~ "immune",

    # assign cancer cells based on condition
    cell_type_df$compartment %in% c("stroma") &
      cell_type_df$has_cnv_score == "CNV" ~ "cancer stroma",
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "CNV" &
      cell_type_df$cell_type == "mesenchymal cell" ~ "blastema",
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "CNV" &
      cell_type_df$cell_type != "mesenchymal cell" ~ "cancer epithelium",
    .default = "unknown"
  ))
```

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "normal", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 0.1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

#### Cancer and normal cells

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in 1:22) {
  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("proportion_cnv_chr", i)))
}
```

@@

```{r}
annotations_table <- cell_type_df |>
  select(
    cell_barcode = barcode,
    scpca_sample_id = sample_id,
    tumor_cell_classification = first.level_annotation,
    cell_type_assignment = second.level_annotation
  ) |>
  mutate(
    # change cancer --> tumor, but keep the other labels
    tumor_cell_classification = ifelse(
      tumor_cell_classification == "cancer", "tumor", tumor_cell_classification
    ),
    cell_type_assignment = str_replace_all(
      cell_type_assignment,
      "cancer ",
      "tumor "
    )
  )

@@

- Combining label transfer and CNV inference we have produced draft annotations for all 40 Wilms tumor samples in SCPCP000006

>>>>

---
title: "Combined annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  predicted.celltype.threshold: 0.75
  cnv_threshold: 0
  testing: 0
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

| first level annotation | second level annotation | selection of the cells | marker genes for validation | CNV validation |
| ---------------------- | ----------------------- | ---------------------- | --------------------------- | --------------- |
| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr  | `VWF`| no CNV |
| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr  | `PTPRC`, `CD163`, `CD68`| no CNV |
| normal                 | kidney                  | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr` | `CDH1`, `PODXL`, `LTL`| no CNV |
| normal                 | stroma                  | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr`| `VIM`| no CNV |
| cancer                 | stroma                  | `compartment == "stroma" & cnv_score > cnv.thr` | `VIM`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | blastema                | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr`  | `CITED1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | epithelial              | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr`  | `CDH1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| unknown                | -                       | the rest of the cells  | -  | -|

@@

The report will be saved in the `notebook` directory.
The final annotation file `SCPCP000006-annotations.tsv` will be saved in the `results` directory.

@@

@@

We simply checked for each chromosome if the cell `has_cnv_chr`.
Would the cell have less than `cnv_threshold` chromosome with CNV, the global `has_cnv_score` will be FALSE.
Would the cell have more than `cnv_threshold` +2 chromosome with CNV, the global `has_cnv_score` will be TRUE.
The default is set to `unknown`.

The `infercnv` method can lead to false positive CNV.
We introduced a flexibility of `+2` over the `cnv_threshold` to reduce the risk of misclassification of normal cells as cancer cells.

Please note that some cancer cell might not have any CNV. 
There is thus a risk of misclassifying cancer cells as normal cells. 
However, we cannot avoid this risk with the data and tools currently available.
This is a limitation of our annotations.

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(has_cnv_score = rowSums(cell_type_df[, grepl("has_cnv_chr", colnames(cell_type_df))])) |>
  mutate(has_cnv_score = case_when(
    has_cnv_score > params$cnv_threshold +2 ~ "CNV",
    has_cnv_score <= params$cnv_threshold ~ "no CNV",
    .default = "unknown"
  ))

```

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
# Define normal cells
# We first pick up the immune and endothelial cells annotated via the label transfer compartments under the condition that the predicted score is above the threshold
cell_type_df <- cell_type_df |>
  mutate(first.level_annotation = case_when(
    # assign normal/cancer based on condition
    compartment %in% c("fetal_nephron", "stroma") & 
      compartment.score > params$predicted.celltype.threshold & 
      has_cnv_score == "no CNV" ~ "normal",

    compartment %in% c("endothelium", "immune") &
      cell_type.score > params$predicted.celltype.threshold ~ "normal",

    compartment %in% c("fetal_nephron", "stroma") & 
      has_cnv_score == "CNV" ~ "cancer",

    .default = "unknown"
  ))

```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = first.level_annotation), shape = 19, size = 1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))

DT::datatable(table(cell_type_df$first.level_annotation, cell_type_df$sample_id))

```

Strikingly, 5 samples show more normal cells than cancer cells:

- `SCPCS000177`
- `SCPCS000180`
- `SCPCS000181`
- `SCPCS000190`
- `SCPCS000197`

These five samples do not have enough immune and endothelium cell to be used as a normal reference in `scripts/06_infercnv.R` and we previously decided to run `scripts/06_infercnv.R` without any reference for these samples.
In that case, the mean expression profile over the sample is taken as the normal reference, biaising the inference of CNV.

For those samples, the annotations based on `infercnv` cannot be trust.
To avoid any confusion in the following steps of the analysis, we decided to force the annotation of the entire samples to `unknown`.

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}

cell_type_df$first.level_annotation[cell_type_df$sample_id %in% c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")] <- "unknown"

DT::datatable(table(cell_type_df$first.level_annotation, cell_type_df$sample_id))

```

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(second.level_annotation = case_when(
    # assign normal cells based on condition
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$compartment.score > params$predicted.celltype.threshold &
      cell_type_df$has_cnv_score == "no CNV" ~ "kidney",

    cell_type_df$compartment %in% c("stroma") &
      cell_type_df$compartment.score > params$predicted.celltype.threshold &
      cell_type_df$has_cnv_score == "no CNV" ~ "normal stroma",

    cell_type_df$compartment %in% c("endothelium") &
      (cell_type_df$compartment.score > params$predicted.celltype.threshold) ~ "endothelium",
    cell_type_df$compartment %in% c("immune") &
      (cell_type_df$compartment.score > params$predicted.celltype.threshold) ~ "immune",

    # assign cancer cells based on condition
    cell_type_df$compartment %in% c("stroma") &
      cell_type_df$has_cnv_score == "CNV" ~ "cancer stroma",
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "CNV" &
      cell_type_df$cell_type == "mesenchymal cell" ~ "blastema",
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "CNV" &
      cell_type_df$cell_type != "mesenchymal cell" ~ "cancer epithelium",
    .default = "unknown"
  ))

cell_type_df$second.level_annotation[cell_type_df$sample_id %in% c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")] <- "unknown"
```

#### `Second.level_annotation` of cancer and normal cells - `umap` reduction

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 0.2, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

#### `Second.level_annotation` of cancer and normal cells without `unknown` - `umap` reduction

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
tmp <- cell_type_df[cell_type_df$second.level_annotation != "unknown",]
ggplot(tmp, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 0.2, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

#### `Second.level_annotation` of cancer cells only - `umap` reduction

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 0.2, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

Per sample, the annotations in the `umap` reduction seem to make sense as we can the three main Wilms tumor components are easily distinguished:

- cancer blastema
- epithelium
- stroma

@@

We look for each `second.level_annotation` and `sample_id`, the proportion of cells that has CNV in each of the chromosome.

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in 1:22) {
  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("has_cnv_chr", i)))
}
```

@@

```{r}
annotations_table <- cell_type_df |>
  dplyr::select(
    cell_barcode = barcode,
    scpca_sample_id = sample_id,
    tumor_cell_classification = first.level_annotation,
    cell_type_assignment = second.level_annotation
  ) |>
  mutate(
    # change cancer --> tumor, but keep the other labels
    tumor_cell_classification = ifelse(
      tumor_cell_classification == "cancer", "tumor", tumor_cell_classification
    ),
    cell_type_assignment = str_replace_all(
      cell_type_assignment,
      "cancer ",
      "tumor "
    )
  )

@@

- Combining label transfer and CNV inference we have produced draft annotations for 35/40 Wilms tumor samples in SCPCP000006.
We decided not to annotate samples for which the `infercnv` couldn't be run with a normal reference, as we don't trust the results in that case and want to avoid misclassification (`SCPCS000177`, `SCPCS000180`, `SCPCS000181`, `SCPCS000190`, `SCPCS000197`).
